\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Scales Data: Application to the Relationship between Reality Beliefs about AI-Generated Faces and Beauty}

### 2025-05-28 paper/manuscript.tex
changes to script

\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, left=2mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, leftrule=.75mm, breakable, rightrule=.15mm, toprule=.15mm, colback=white, opacityback=0]

This preprint is a non-peer-reviewed work from the
\href{https://realitybending.github.io/}{\textbf{Reality Bending Lab}}.
\begin{center}
\includegraphics[width=0.2\textwidth,height=\textheight]{manuscript_files/mediabag/ReBeL_LogoOnly_hu114.png}
\end{center}

@@

Despite significant advancements in psychological science following the
replication crisis (\citeproc{ref-open2015estimating}{Collaboration,
2015}), its progress is still hindered by its sub-optimal (or
inappropriate) usage of statistical tools
(\citeproc{ref-makowski2023we}{Makowski \& Waggoner, 2023}). A prevalent
issue is the continued reliance on linear models that assume normally
distributed (Gaussian) data - as this assumption often does not hold
true for many types of psychological outcomes. For instance, reaction
times typically exhibit skewed distributions, choices can be represented
as binary variables, and count data consists of strictly positive
integers. Applying models that presume normality and model the ``mean''
of the outcome variable can lead to misinterpretation and potentially
misleading conclusions when applied indiscriminately. It is thus
important that psychologists use models that can best describe the data
they collect, to fully exploit them and bring more nuance and accuracy
to their conclusions.

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{./figures/fig1.png}

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{./figures/fig2.png}

@@

Another unknown is whether these beliefs are fully stimuli-driven or
whether there are some stable individual tendencies that are at stake.

@@

The sample is the same as in (\citeproc{ref-makowski2025too}{Makowski et
al., 2025}). We included all heterosexual male and female participants,
for a final sample of 125 participants (Mean age = 28.6, SD = 9.2,
range: {[}19, 66{]}; Sex: 44.8\% females). We included only sexually
relevant stimuli for men and women (i.e., all 89 female faces for men
and 20 male faces for women).

@@

In the first phase, participants viewed 109 neutral-expression
photographs of faces (random order, display time of 500 ms) from the
American Multiracial Face Database
(\citeproc{ref-chen2021broadening}{Chen et al., 2021}). After each
image, participants rated the face on trustworthiness, familiarity,
attractiveness, and beauty using visual analog scales. In the second
phase, participants were informed that ``about half of the previously
seen images were AI-generated''. The same faces were presented again in
a new random order (same display time), followed by ratings of
``reality'' (whether they believed the image was AI-generated - left
anchor - or a real photograph - right anchor). Makowski et al.
(\citeproc{ref-makowski2025too}{2025}) showed that most stimuli yielded
highly variable reality ratings, suggesting that idiosyncratic cognitive
processes were at play.

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a
Beta-Gate model, and the CHOCO model. For all models and each parameter,
the full formula was entered:
\(Real\sim Sex + (1 | Participant) + (1 | Item)\) (with Sex as the main
predictor and participants and items entered as random intercepts). The
models were run using \emph{brms}
(\citeproc{ref-burkner2017brms}{Bürkner, 2017}) R package, and analyzed
using the \emph{easystats} collection of packages {[}\textbf{REFS}{]}.
To maximize the comparability across models We used the default priors
(uniform) for all models, and we ran 16 chains of 1400 iterations each
on the University of Sussex High-Performance Computing (HPC) cluster.

@@

For the population-level effects, we will consider significant and
report (using the median of the posterior distribution) effects for
which the 95\% Credible Interval (CI) does not include zero {[}and when
the probability of direction \emph{pd} is \textgreater{}
\textasciitilde97\%; Makowski et al.
(\citeproc{ref-makowski2019indices}{2019}){]}. For the individual-level
parameters (i.e., the random intercepts of each parameter for each
participant and each item), we will first analyze their reliability
using the Variance-Over-Uncertainty Ratio index (\emph{D-vour}). This
index, implemented in the \emph{performance} package
(\citeproc{ref-ludecke2021performance}{Lüdecke et al., 2021}), is
inspired by recent work on mixed models reliability
(\citeproc{ref-rouder2024hierarchical}{Rouder \& Mehrvarz, 2024};
\citeproc{ref-williams2021beneath}{Williams et al., 2021}), and
corresponds to the normalized ratio of observed variability to
uncertainty in random effect estimates, defined as:

@@

\includegraphics[width=1\textwidth,height=\textheight]{./figures/fig3.png}

@@

The models did converge without divergent transitions, and the effective
sample size was sufficient for all parameters (all
\(n_{\text{eff}} > 1000\)). The difference in predictive accuracy, as
indexed by Expected Log Predictive Density (ELPD-WAIC), suggests that
the \emph{CHOCO} is the best model (\(ELPD = -138.45\)), followed by
\emph{Beta-Gate} (\(\Delta_{ELPD} = -1875.31 \pm 61.18, p < .001)\) and
\emph{ZOIB} (\(\Delta_{ELPD} = -1916.11 \pm 61.75, p < .001\)). See
Figure~\ref{fig-three} for the posterior predictive checks, showing that
only the CHOCO model managed to capture the bimodal distribution of
data.

@@

The ZOIB model suggested that women had higher mean (\emph{mu}
parameter) scores of reality beliefs
(\(\Delta_{\text{Female}} = 0.16,~95\%~ CI~[0.01, 0.32], pd = 98.10\%\)),
less extreme values (\emph{zoi} parameter; STATS) but more ones relative
to zeros (\emph{coi} parameter; STATS).

@@

Across all these models, no effect of Sex on the prediction parameter
was observed.

@@

\includegraphics[width=1\textwidth,height=\textheight]{./figures/fig4.png}

@@

The ZOIB model estimated reliable variability in the participant's
\emph{phi} parameter (D-vour = 0.88) and \emph{zoi} parameter (D-vour =
0.84), as well as in the \emph{mu} parameter related to individual items
(D-vour = 0.82). Moderate reliability was also observed for items in the
\emph{coi} parameter (D-vour = 0.71) and for participants in the
\emph{mu} parameter (D-vour = 0.69). The Beta-Gate model yielded similar
results: a high reliability of participant's \emph{phi} parameter
(D-vour = 0.88), \emph{pex} parameter (D-vour = 0.84). The \emph{mu}
parameter's variability was reliably captured for items (D-vour = 0.85)
and moderately for participants (D-vour = 0.72). Of note that the
parameters of Beta-Gate and ZOIB correlate almost perfectly, underlining
their empirical similarity despite a different parametrization.

The CHOCO model yielded reliable estimates for all parameters except
\emph{bex} for participants (\emph{confright} D-vour = 0.94,
\emph{confleft} D-vour = 0.91, \emph{pex} D-vour = 0.80, \emph{p} D-vour
= 0.73, \emph{precright} D-vour = 0.79, \emph{precleft} D-vour = 0.73).
Item's variability was primarily reflected through the \emph{p}
parameter (D-vour = 0.86).

Finally, the empirical average correlated the most strongly with CHOCO's
\emph{p} (r = .94), rather than ZOIB's \emph{mu} (r = .74) or
Beta-Gate's \emph{mu} (r = .82). The empirical overall confidence was
the strongest correlated with CHOCO's \emph{confright} (r = .91),
followed by ZOIB's \emph{phi} (r = -.86), Beta-Gate's phi (r = -.83),
and other CHOCO's parameters. The empirical proportion of ``right''
answer \emph{p} correlated the strongest with CHOCO's \emph{p} (r =
0.94), followed by Beta-Gate's \emph{mu} (r = .82) and ZOIB's \emph{mu}
(r = .77). The empirical \emph{pex} correlated the strongest with ZOIB's
\emph{zoi} (r = .90), and Beta-Gate's \emph{pex} (r = .90), and CHOCO's
\emph{pex} (r = .88).

@@

The explosion of accessibility of state-of-the-art AI tools has made it
effortless to generate realistic images, including human faces that are
often indistinguishable from real ones \textbf{{[}REF{]}}. These
synthetic visuals are flooding the cyberspace across various domains,
such as art, advertising and entertainment, to education and
information. This technological advancement carries an important
potential for misuse, such as in disinformation campaigns, scams (e.g.,
AI-bots, identity theft), and abuse. The democratization of such
technology raises pressing concerns about the value of authenticity and
the potential erosion of media trust in our increasingly
\emph{post-truth} society.

In this evolving landscape, understanding the cognitive mechanisms that
underpin our judgments of reality becomes paramount. Despite the
increasing prevalence of digitally altered or AI-generated content,
humans still rely on certain heuristics to assess authenticity. One such
heuristic might be facial attractiveness. Attractiveness appraisals are
known to be automatic and unconscious (\citeproc{ref-hou2023neural}{Hou
et al., 2023}; \citeproc{ref-hung2016unconscious}{Hung et al., 2016};
\citeproc{ref-luo2019robust}{Luo et al., 2019}), and carry strong
real-life consequences, as demonstrated by the large body of literature
on the ``beauty premium'' (\citeproc{ref-gulati2024beautiful}{Gulati et
al., 2024}, \textbf{MORE}; \citeproc{ref-little2021facial}{Little,
2021})\footnote{As an eloquent example, Monk Jr et al.
  (\citeproc{ref-monk2021beholding}{2021}) reported that the the
  magnitude of the earnings disparities among white women along the
  perceived attractiveness continuum exceeds in magnitude the canonical
  black-white race gap in a large representative US sample.}

\textbf{TODO:} However, it is important to allow for the possibility of
a U-shaped relationship. For instance, Peng et al.~(2020) reported that
both attractive and unattractive people sold more (underlining a
``plainness penalty'').

Although its role as a potential modulator of reality beliefs remains
underexplored, Makowski et al\ldots{} \textbf{TODO: detail findings of
that study + other studies}.

@@

The first sample includes the same participants as study 1 (see above).
\textbf{TODO} for sample 2.

@@

For the second sample, the procedure was relatively similar with a few
key differences. \textbf{TODO} Describe fake Face 2. The main difference
is that the subjective ratings were done on a X-point Likert scale,
rather than analog, with a mid-point value.

@@

In sample 1, the traditional approach suggested a significant linear
relationship between the mean level of ``Reality'' and attractiveness
(\(\beta_{poly1}=3.63, ~95\%~CI [2.71, 4.56], pd = ...\%\)) for males
only. The CHOCO model revealed that attractiveness had a significantly
positive linear relationship with the probability \emph{p} of judging
faces as real (STATS), but a quadratic relationship with the confidence
in real judgments (\emph{confright}; STATS). Attractiveness was also
associated with less confidence in fake judgments (\emph{confleft};
STATS), and related linearly with a stronger bias towards extreme Real
responses (\emph{bex}; STATS). These effects were present for males
only, and no relationship was found for females. The reliability of the
effect of attractiveness (the random slopes) was very low for all
parameters (D-vour \textless{} 0.1).

The effect of Beauty was the same as above, with some additional
evidence in favour of an effect in women, in particular for a linear
relationship with \emph{p} for females as well
(\(\beta_{poly1}=7.02, ~95\%~CI [-0.90, 14.97], pd = 95.91\%\)), as well
as a quadradic relationship with the confidence in fake judgments
(\emph{confleft}; STATS).

@@

@@

We would like to thank the dissertation students from the University of
Sussex for their help in data collection.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-burkner2017brms}
Bürkner, P.-C. (2017). Brms: An r package for bayesian multilevel models
using stan. \emph{Journal of Statistical Software}, \emph{80}, 1--28.

@@

@@

@@

@@

@@

@@

@@

>>>>

\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

@@

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, bottomrule=.15mm, colback=white, colframe=quarto-callout-note-color-frame, opacityback=0, breakable, arc=.35mm, leftrule=.75mm, toprule=.15mm, left=2mm]

This preprint is a non-peer-reviewed work from the
\href{https://realitybending.github.io/}{\textbf{Reality Bending Lab}}.
\begin{center}
\includegraphics[width=0.2\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/ReBeL_LogoOnly_hu114.png}
\end{center}

@@

Despite significant advancements in psychological science following the
replication crisis (\citeproc{ref-open2015estimating}{Collaboration,
2015}), its progress is still hindered by its sub-optimal (or
inappropriate) usage of statistical tools
(\citeproc{ref-makowski2023we}{Makowski \& Waggoner, 2023}). A prevalent
issue is the continued reliance on linear models that assume normally
distributed (Gaussian) data - as this assumption often does not hold
true for many types of psychological outcomes. For instance, reaction
times typically exhibit skewed distributions, choices can be represented
as binary variables, and count data consists of strictly positive
integers. Applying models that presume normality and model the ``mean''
of the outcome variable can lead to misinterpretation and potentially
misleading conclusions when applied indiscriminately. It is thus
important that psychologists use models that can best describe (or
generate) the data they collect, to fully exploit them and bring more
nuance and accuracy to their conclusions.

@@

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{./figures/fig1.png}

@@

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{./figures/fig2.png}

@@

In a post-truth era (\citeproc{ref-lewandowsky2017beyond}{Lewandowsky et
al., 2017}), where near-perfect simulations are increasingly feasible
due to rapid technological advancements, distinguishing between what is
real and what is fake poses a growing challenge {[}\textbf{REF}{]}. As
these technologies advance and physical cues become indistinguishable,
individuals are likely to rely more heavily on alternative sources of
information -such as contextual cues and cognitive heuristic -when
making reality judgments {[}i.e., simulation monitoring; Makowski
(\citeproc{ref-makowski2018cognitive}{2018}), Makowski, Sperduti, et al.
(\citeproc{ref-makowski2019phenomenal}{2019}){]}. This reliance is
particularly pronounced under conditions of high ambiguity, where
processing demands are higher and the atomization and
decontextualization of information (especially prevalent in online
environments) further complicate the discernment of authenticity.

Another unknown is whether these beliefs are fully stimuli-driven or
whether there are some stable individual tendencies that are at stake.
For instance, in the absence of clear contextual information,
individual-level characteristics may increasingly guide beliefs about
reality. Traits such as narcissism have been shown to influence such
judgments (\citeproc{ref-makowski2025too}{Makowski et al., 2025}).
Moreover, it has been proposed that the interaction between stimulus
characteritics and individual differences plays a crucial role in
shaping reality beliefs. For example, self-reported arousal has been
found to predict the likelihood of perceiving an image as real
(\citeproc{ref-azevedo2020body}{Azevedo et al., 2020}), while judging a
stimulus as fake has been associated with emotional downregulation
(\citeproc{ref-makowski2019phenomenal}{Makowski, Sperduti, et al.,
2019}). {[}\textbf{should probably add other examples}{]}

These findings underscore that what drives beliefs and judgments of
reality remains unclear. Yet, understanding these mechanisms is
increasingly vital given the proliferation of AI-generated content and
the growing risk of misjudging the authenticity of stimuli, particularly
in contexts with significant societal consequences (e.g., the influence
of misinformation on democratic processes).

@@

The sample is the same as in (\citeproc{ref-makowski2025too}{Makowski et
al., 2025}). We included all heterosexual and bisexual (these two groups
did not seem to differ based on preliminary analyses and were thus
grouped to maximize power) male and female participants, for a final
sample of 141 participants (Mean age = 28.4, SD = 9.0, range: {[}19,
66{]}; Sex: 47.5\% females). For each participant, we included only
stimuli of the opposite gender (i.e., all 89 female faces for men and 20
male faces for women).

@@

In the first phase, participants viewed 109 neutral-expression
photographs of faces (random order, display time of 500 ms) from the
American Multiracial Face Database
(\citeproc{ref-chen2021broadening}{Chen et al., 2021}). After each
image, participants rated the face on trustworthiness, familiarity,
attractiveness, and beauty using visual analog scales. In the second
phase, participants were informed that ``about half of the previously
seen images were AI-generated''. The same faces were presented again in
a new random order (same display time), followed by ratings of
``reality'' (whether they believed the image was fake - left anchor - or
a real - right anchor). Makowski et al.
(\citeproc{ref-makowski2025too}{2025}) showed that most stimuli yielded
highly variable reality ratings, suggesting that idiosyncratic cognitive
processes were at play.

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a
Beta-Gate model, and the CHOCO model. For all models and each parameter,
the full formula was entered:
\(Real\sim Sex + (1 | Participant) + (1 | Item)\) (with Sex as the main
predictor and participants and items entered as random intercepts). The
models were run using \emph{brms}
(\citeproc{ref-burkner2017brms}{Bürkner, 2017}) R package, and analyzed
using the \emph{easystats} collection of packages
(\citeproc{ref-easystats2022}{Lüdecke et al., 2022}). To maximize the
comparability across models We used the default priors (uniform) for all
models, and we ran 16 chains of 1400 iterations each on the University
of Sussex High-Performance Computing (HPC) cluster.

@@

For the population-level effects, we will consider significant and
report (using the median of the posterior distribution) effects for
which the 95\% Credible Interval (CI) does not include zero (and when
the probability of direction \emph{pd} is \textgreater{}
\textasciitilde97\%, \citeproc{ref-makowski2019indices}{Makowski,
Ben-Shachar, et al., 2019}). For the individual-level parameters (i.e.,
the random intercepts of each parameter for each participant and each
item), we will first analyze their reliability using the
Variance-Over-Uncertainty Ratio index (\emph{D-vour}). This index,
implemented in the \emph{performance} package
(\citeproc{ref-ludecke2021performance}{Lüdecke et al., 2021}), is
inspired by recent work on mixed models reliability
(\citeproc{ref-rouder2024hierarchical}{Rouder \& Mehrvarz, 2024};
\citeproc{ref-williams2021beneath}{Williams et al., 2021}), and
corresponds to the normalized ratio of observed variability to
uncertainty in random effect estimates, defined as:

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig3.png}

@@

The models did converge without divergent transitions, and the effective
sample size was sufficient for all parameters (all
\(n_{\text{eff}} > 1000\)). The difference in predictive accuracy, as
indexed by Expected Log Predictive Density (ELPD-WAIC), suggests that
the \emph{CHOCO} is the best model (\(ELPD = -203.54\)), followed by
\emph{Beta-Gate} (\(\Delta_{ELPD} = -1794.57 \pm 63.12, p < .001)\) and
\emph{ZOIB} (\(\Delta_{ELPD} = -1833.59 \pm 63.52, p < .001\)). See
Figure~\ref{fig-three} for the posterior predictive checks, showing that
only the CHOCO model managed to capture the bimodal distribution of
data.

@@

The ZOIB model suggested that women had higher mean (\emph{mu}
parameter) scores of reality beliefs
(\(\Delta_{\text{Female}} = 0.20,~95\%~ CI~[0.03, 0.37], pd = 98.77\%\)),
less extreme values (\emph{zoi} parameter; STATS) but more ones relative
to zeros (\emph{coi} parameter; STATS).

@@

Across all these models, no effect of Sex on the precision parameter was
observed.

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig4.png}

@@

The ZOIB model estimated reliable variability in the participant's
\emph{phi} parameter (D-vour = 0.88) and \emph{zoi} parameter (D-vour =
0.85), as well as in the \emph{mu} parameter related to individual items
(D-vour = 0.82). Moderate reliability was also observed for items in the
\emph{coi} parameter (D-vour = 0.71) and for participants in the
\emph{mu} parameter (D-vour = 0.69). The Beta-Gate model yielded similar
results: a high reliability of participant's \emph{phi} parameter
(D-vour = 0.88), \emph{pex} parameter (D-vour = 0.85). The \emph{mu}
parameter's variability was reliably captured for items (D-vour = 0.85)
and moderately for participants (D-vour = 0.72).

The CHOCO model yielded reliable estimates for all parameters except
\emph{bex} for participants (\emph{confright} D-vour = 0.94,
\emph{confleft} D-vour = 0.91, \emph{pex} D-vour = 0.79, \emph{p} D-vour
= 0.73, \emph{precright} D-vour = 0.77, \emph{precleft} D-vour = 0.67).
Item's variability was primarily reflected through the \emph{p}
parameter (D-vour = 0.86).

Finally, the empirical average correlated the most strongly with CHOCO's
\emph{p} (r = .86), rather than ZOIB's \emph{mu} (r = .77) or
Beta-Gate's \emph{mu} (r = .82). The empirical overall confidence was
the strongest correlated with CHOCO's \emph{confright} (r = .91),
followed by ZOIB's \emph{phi} (r = -.86), Beta-Gate's phi (r = -.83),
and other CHOCO's parameters. The empirical proportion of ``right''
answer \emph{p} correlated the strongest with CHOCO's \emph{p} (r =
0.94), followed by Beta-Gate's \emph{mu} (r = .82) and ZOIB's \emph{mu}
(r = .77). The empirical \emph{pex} correlated the strongest with ZOIB's
\emph{zoi} (r = .90), and Beta-Gate's \emph{pex} (r = .90), and CHOCO's
\emph{pex} (r = .88). Of note that the parameters of Beta-Gate and ZOIB
correlate almost perfectly, underlining their empirical similarity
despite a different parametrization.

@@

The explosion of accessibility of state-of-the-art AI tools has made it
effortless to generate realistic images, including human faces that are
often indistinguishable from real ones
(\citeproc{ref-bozkir2024can}{Bozkir et al., 2024};
\citeproc{ref-miller2023ai}{Miller et al., 2023};
\citeproc{ref-nightingale2022ai}{Nightingale \& Farid, 2022}). These
synthetic visuals are flooding the cyberspace across various domains,
such as art, advertising and entertainment, to education and
information. This technological advancement carries an important
potential for misuse, such as in disinformation campaigns, scams (e.g.,
AI-bots, identity theft), and abuse. The democratization of such
technology raises pressing concerns about the value of authenticity and
the potential erosion of media trust in our increasingly
\emph{post-truth} society.

In this evolving landscape, understanding the cognitive mechanisms that
underpin our judgments of reality becomes paramount. Despite the
increasing prevalence of digitally altered or AI-generated content,
humans still rely on certain heuristics to assess authenticity. One such
heuristic might be facial attractiveness. Attractiveness appraisals are
known to be automatic and unconscious (\citeproc{ref-hou2023neural}{Hou
et al., 2023}; \citeproc{ref-hung2016unconscious}{Hung et al., 2016};
\citeproc{ref-luo2019robust}{Luo et al., 2019}), and carry strong
real-life consequences, as demonstrated by the large body of literature
on the ``beauty premium'' (\citeproc{ref-gulati2024beautiful}{Gulati et
al., 2024}; \citeproc{ref-kukkonen2024beauty}{Kukkonen et al., 2024};
\citeproc{ref-little2021facial}{Little, 2021};
\citeproc{ref-pandey2021face}{Pandey \& Zayas, 2021})\footnote{As an
  eloquent example, Monk Jr et al.
  (\citeproc{ref-monk2021beholding}{2021}) reported in a large
  representative US sample that the magnitude of earnings disparities
  among white women along the perceived attractiveness continuum exceeds
  in magnitude the canonical black-white race gap.}

\textbf{TODO:} However, it is important to allow for the possibility of
a U-shaped relationship. For instance, Peng et al.
(\citeproc{ref-peng2020faces}{2020}) reported that both attractive and
unattractive people sold more products (underlining a ``plainness
penalty'').

Although its role as a potential modulator of reality beliefs remains
underexplored, Makowski et al. (\citeproc{ref-makowski2025too}{2025})
found significant associations between participants' realness ratings
and facial salience. Female participants judged both highly attractive
and highly unattractive faces as more likely to be real, whereas for
male participants, this effect was limited to highly attractive faces.
These findings offer a complementary perspective to those of Miller et
al. (\citeproc{ref-miller2023ai}{2023}), who reported that AI-generated
faces tend to be rated as more attractive, and that participants used
attractiveness as a distinguishing cue between real and AI-generated
faces. Together, these results highlight a possible bidirectional and
context-dependent role of attractiveness in shaping reality judgments.

@@

The first sample includes the same participants as study 1 (see above).

\textbf{TODO} for sample 2.

The final sample included 189 participants (Mean age = 28.4, SD = 14.0,
range: {[}18, 69{]}; Sex: 76.2\% females).

@@

For the second sample, the procedure was relatively similar with a few
key differences. Most importantly, subjective ratings were collected
using a 7-point Likert scale ranging from 0 to 6, rather than a visual
analog scale, which included a clear midpoint. Additionally, the
familiarity rating was removed from the set of facial evaluations. In
the second phase, realness judgments were also recorded on a 7-point
Likert scale, this time ranging from -3 to +3, again with a defined
midpoint, replacing the visual analog format used in the first study.
For this scale, the endpoints were labeled `AI-generated' and
`Photograph', replacing the `Fake' and `Real' anchors used in Study 1.

There were also modifications to the instructions provided to
participants. In Phase 1, an additional prompt was introduced, stating
that the study was conducted in partnership with a young AI startup.
This was intended to strengthen participants' belief that the faces they
were shown could plausibly be either real or AI-generated. Additionally,
participants were explicitly asked to respond based on their immediate,
first impressions---an instruction not included in the first study.

In Phase 2, participants were not informed about the proportion of
AI-generated versus real faces encountered during Phase 1. This omission
was deliberate, aimed at reducing demand characteristics and preventing
participants from mechanically selecting half the faces as real and half
as AI-generated. Instead, they were encouraged to base their responses
solely on their level of confidence regarding the origin of each image.

@@

\subsubsection{Sample 1}\label{sample-1}

In sample 1, the traditional approach suggested a significant linear
relationship between the mean level of ``Reality'' and attractiveness
(\(\beta_{poly1}=3.42, ~95\%~CI [2.50, 4.34], p < .001\%\)) for males
only. The CHOCO model revealed that attractiveness had a significantly
positive linear relationship with the probability \emph{p} of judging
faces as real (STATS), but a quadratic relationship with the confidence
in real judgments (\emph{confright}; STATS). Attractiveness was also
associated with less confidence in fake judgments (\emph{confleft};
STATS), a quadratic relationship with left precision (\emph{precleft};
STATS), and related linearly with a stronger bias towards extreme Real
responses (\emph{bex}; STATS). These effects were present for males
only, and no relationship was found for females. The reliability of the
effect of attractiveness (the random slopes) was very low for all
parameters (D-vour \textless{} 0.01). The effect of Beauty was the same
as of attractiveness.

\subsubsection{Sample 2}\label{sample-2}

@@

This study introduces a new statistical model to analyze bimodal data,
which can be observed for instance when using subjective rating scales
in which the two sides correspond to a different ``choice'' (e.g.,
``false'' vs.~``true'', ``real'' vs ``fake''). The Choice-Confidence
(CHOCO) model conceptualizes the data as a mixture of two distributions,
one for each choice, and models the probability of choosing one or the
other as well as the degree of confidence in each choice. We validated
this model on real-life data by appling it to the reality beliefs of
Makowski et al. (\citeproc{ref-makowski2025too}{2025}), and showed that
it was able to capture the bimodal nature of the data and provides a
more accurate representation of the underlying processes than other
alternative approaches, including Zero-or-One inflated Beta (ZOIB)
models.

@@

We would like to thank the dissertation students from the University of
Sussex for their help in data collection. DM would also like to thank
\ldots{} for the motivation provided to write this paper.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-azevedo2020body}
Azevedo, R., Tucciarelli, R., De Beukelaer, S., Ambroziak, K., Jones,
I., \& Tsakiris, M. (2020). \emph{A body of evidence:`feeling in
seeing'predicts realness judgments for photojournalistic images.}

\bibitem[\citeproctext]{ref-bozkir2024can}
Bozkir, E., Riedmiller, C., Skodras, A. N., Kasneci, G., \& Kasneci, E.
(2024). Can you tell real from fake face images? Perception of
computer-generated faces by humans. \emph{ACM Transactions on Applied
Perception}, \emph{22}(2), 1--23.

\bibitem[\citeproctext]{ref-burkner2017brms}
Bürkner, P.-C. (2017). Brms: An r package for bayesian multilevel models
using stan. \emph{Journal of Statistical Software}, \emph{80}, 1--28.

@@

\bibitem[\citeproctext]{ref-kukkonen2024beauty}
Kukkonen, I., Pajunen, T., Sarpila, O., \& Åberg, E. (2024). Is
beauty-based inequality gendered? A systematic review of gender
differences in socioeconomic outcomes of physical attractiveness in
labor markets. \emph{European Societies}, \emph{26}(1), 117--148.

\bibitem[\citeproctext]{ref-lewandowsky2017beyond}
Lewandowsky, S., Ecker, U. K., \& Cook, J. (2017). Beyond
misinformation: Understanding and coping with the {``post-truth''} era.
\emph{Journal of Applied Research in Memory and Cognition}, \emph{6}(4),
353--369.

@@

\bibitem[\citeproctext]{ref-easystats2022}
Lüdecke, D., Ben-Shachar, M. S., Patil, I., Wiernik, B. M., Bacher, E.,
Thériault, R., \& Makowski, D. (2022). Easystats: Framework for easy
statistical modeling, visualization, and reporting. \emph{CRAN}.
\url{https://doi.org/10.32614/CRAN.package.easystats}

@@

\bibitem[\citeproctext]{ref-makowski2018cognitive}
Makowski, D. (2018). \emph{Cognitive neuropsychology of implicit emotion
regulation through fictional reappraisal} {[}PhD thesis{]}. Sorbonne
Paris Cit{é}.

@@

\bibitem[\citeproctext]{ref-makowski2019phenomenal}
Makowski, D., Sperduti, M., Pelletier, J., Blondé, P., La Corte, V.,
Arcangeli, M., Zalla, T., Lemaire, S., Dokic, J., Nicolas, S., et al.
(2019). Phenomenal, bodily and brain correlates of fictional reappraisal
as an implicit emotion regulation strategy. \emph{Cognitive, Affective,
\& Behavioral Neuroscience}, \emph{19}, 877--897.

@@

\bibitem[\citeproctext]{ref-miller2023ai}
Miller, E. J., Steward, B. A., Witkower, Z., Sutherland, C. A.,
Krumhuber, E. G., \& Dawel, A. (2023). AI hyperrealism: Why AI faces are
perceived as more real than human ones. \emph{Psychological Science},
\emph{34}(12), 1390--1403.

@@

\bibitem[\citeproctext]{ref-nightingale2022ai}
Nightingale, S. J., \& Farid, H. (2022). AI-synthesized faces are
indistinguishable from real faces and more trustworthy.
\emph{Proceedings of the National Academy of Sciences}, \emph{119}(8),
e2120481119.

@@

\bibitem[\citeproctext]{ref-pandey2021face}
Pandey, G., \& Zayas, V. (2021). What is a face worth? Facial
attractiveness biases experience-based monetary decision-making.
\emph{British Journal of Psychology}, \emph{112}(4), 934--963.

\bibitem[\citeproctext]{ref-peng2020faces}
Peng, L., Cui, G., Chung, Y., \& Zheng, W. (2020). The faces of success:
Beauty and ugliness premiums in e-commerce platforms. \emph{Journal of
Marketing}, \emph{84}(4), 67--85.
